---
title: "PHP2514 - Homework 1"
author: "Jess Kaminsky"
date: "February 14, 2018"
output: html_document
---

*Question 1*
```{r}
wine <- read.csv('Wine.csv')
attach(wine)
```

Do these data suggest that heart disease death rates are associated with average wine consumption? If so, how can that be described?

In assessing the relationship between heart disease death rates and average wine consumptions, we will begin by performing simple exploratory data analysis on the data. There are 18 countries represented in this data set and we have average rates of wine consumption and heart disease mortality rates for each country. The purpose of this analysis is to determine if heart disease mortality is associated with average wine consumption and the nature of that relationship, if it exists.
```{r}
summary(wine)
```
```{r}
plot(WINE, MORTALITY, xlab = "Wine Consumption", ylab = "Heart Disease Mortality Rate")
```

```{r}
cor(WINE, MORTALITY)
model1 <- lm(MORTALITY ~ WINE, wine)
summary(model1)
plot(model1)
```

```{r}
cor(log(WINE), log(MORTALITY))
plot(log(WINE), log(MORTALITY), xlab = "Log of Wine Consumption", ylab = "Log of Mortality Rate")
model2 <- log(MORTALITY) ~ log(WINE), wine)
lm(model2)
plot(model2)
```

**Do any countries have substantially higher or lower death rates than others with similar wine consumption rates?**

#look for a significant difference in mean death rate within each level group?
```{r}
### add a column for wine consumption level - a categorical variable
wine$level[1:6] <- "low"
wine$level[7:12] <- "medium"
wine$level[13:18] <- "high"
```

*Question 2*

Part A:
```{r}
flowers <- read.csv('flowers.csv')
flowers[,4] <- rep(c(1,2))
names(flowers)[4] <- "REPLICATE"
flowers$REPLICATE <- as.factor(flowers$REPLICATE)
```

Part B:
```{r}
flowers[,5] <- NA
names(flowers)[5] <- "INTENS_CAT"
for (i in 1:nrow(flowers)) {
  if(flowers$INTENS[i] < 275) {
    flowers$INTENS_CAT[i] <- 1
  }
  else if (flowers$INTENS[i] < 400) {
    flowers$INTENS_CAT[i] <- 2
  }
  else if (flowers$INTENS[i] < 525) {
    flowers$INTENS_CAT[i] <- 3
  }
  else if (flowers$INTENS[i] < 650) {
    flowers$INTENS_CAT[i] <- 4
  }
  else if (flowers$INTENS[i] < 775) {
    flowers$INTENS_CAT[i] <- 5
  }
  else { 
    flowers$INTENS_CAT[i] <- 6
  }
}
flowers$INTENS_CAT <- as.factor(flowers$INTENS_CAT)
```

Part C:
The research questions are: *What are effects of intensity and timing? Is there an interaction between the two factors?*

Part D:
First create an analysis of variance using timing and the categorical form of the light intensity variable. Determine if there is an effect of each factor.
```{r}
summary(aov(FLOWERS ~ TIME + INTENS_CAT, flowers))
m1 <- lm(FLOWERS ~ TIME + INTENS_CAT, flowers)
```

Part E:
Then create an interaction between light intensity and timing by multiplying the two variables and test for the presence of an interaction.
```{r}
summary(aov(FLOWERS ~ TIME + INTENS_CAT + TIME*INTENS_CAT, flowers))
m2 <- lm(FLOWERS ~ TIME + INTENS_CAT + TIME*INTENS_CAT, flowers)
summary(m2)
```

Part F:
Now repeat the process but using light intensity as a continuous variable.
```{r}
summary(aov(FLOWERS ~ TIME + INTENS, flowers))
m3 <- lm(FLOWERS ~ TIME + INTENS, flowers)
summary(aov(FLOWERS ~ TIME + INTENS + TIME*INTENS, flowers))
m4 <- lm(FLOWERS ~ TIME + INTENS + TIME*INTENS, flowers)
```

Part G:
Then perform F-tests to compare the four model you have created (light as continuous and categorical with and without the interaction)
```{r}
anova(m1, m2, m3, m4)
```

Part H:
Predict the number of flowers grown at each combination of light and timing for each of the four models.
```{r}
m1_predict <- predict(m1)
m2_predict <- predict(m2)
m3_predict <- predict(m3)
m4_predict <- predict(m4)

predicted_values <- cbind(time = flowers$TIME, intensity = flowers$INTENS, 
                          intensity_cat = flowers$INTENS_CAT,
                          m1 = m1_predict, m2 = m2_predict, m3 = m3_predict,
                          m4 = m4_predict)
```

Part I:
Compare each prediction to the observed number of flowers and calculate the difference (observed – predicted). This is the residual. Calculate the residual mean squared error for each model by adding the squared residuals together and dividing by the number of residual degrees of freedom. This should equal the mean squared error in each ANOVA table.
```{r}
##calculate residual values
residual_values <- cbind(m1 = flowers$FLOWERS - m1_predict, m2 = flowers$FLOWERS - m2_predict,
                         m3 = flowers$FLOWERS - m3_predict, m4 = flowers$FLOWERS - m4_predict)

##calculate MSE manually and show that it is equal to the MSE from the anova table
MSE_m1 <- sum(residual_values[,1] * residual_values[,1]) / m1$df.residual
ANOVA_MSE_1 <- summary(aov(m1))[[1]] $ `Mean Sq`[3]

MSE_m2 <- sum(residual_values[,2] * residual_values[,2]) / m2$df.residual
ANOVA_MSE_2 <- summary(aov(m2))[[1]] $ `Mean Sq`[4]

MSE_m3 <- sum(residual_values[,3] * residual_values[,3]) / m3$df.residual
ANOVA_MSE_3 <- summary(aov(m3))[[1]] $ `Mean Sq`[3]

MSE_m4 <- sum(residual_values[,4] * residual_values[,4]) / m4$df.residual
ANOVA_MSE_4 <- summary(aov(m4))[[1]] $ `Mean Sq`[4]
```

Part J:
Now plot the residuals vs. the predicted for each model and see if there are any patterns. If you see any, what might you do to remove them?
```{r}
plot(residual_values[,1], predicted_values[,4])
plot(residual_values[,2], predicted_values[,5])
plot(residual_values[,3], predicted_values[,6])
plot(residual_values[,4], predicted_values[,7])
```

Part K:
Finally, take the model you think describes the data the best and write a short report for your grandmother who would like to grow these flowers carefully explaining to her how she should best grow them and why. Note that your grandmother is curious about how much changes in light and timing might affect her flowers and how sensitive her results will be to the settings she makes.

